Forecasting biodiversity in breeding birds using best practices

### 2017-12-08 manuscript/reviewer_reply.Rmd

*The reviewers and I agree that this manuscript is interesting, timely and well-written. I will add a few of my own observations here. First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.*

>Thank you. We have removed the right justification and hyphenation.

*Next, in your figure 1, which forms the basis for comparison of model accuracy over time, I really expected you to choose data WITH a temporal pattern. After all, it looks to me like you have a time series in which species richness simply varies around a more or less stable average value, and the variation can be due to nothing (random), observer bias, and any number of species population trends (and combinations thereof). Thus, you have a situation of years of no trend, and use those data to predict no trend into the future? I would have thought you would choose data with a definite time trend that was already potentially correlated with, say, climate change, and then use hindcasting and forecasting to test whether the models accurately predict a trend into the "future"! To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*

>We chose the North American Breeding Bird Survey data because it is one of the largest scale most long running spatiotemporal datasets available, thus making this kind of quantitative assessment of both spatial and temporally focused forecasting methods possible. We agree that the relative stability of these time-series is important, which is why we dedicated a large discussion paragraph to this point in the initial submission. To further emphasize this point and to highlight the idea that assessing these methods on selectively chosen data exhibiting significant trends is interesting we have added an additional paragraph to the discussion addressing this point directly and recommending it for future research.

@@

#### Comments for the author

*In this paper, the authors use the incredibly rich data from the Breeding Bird Survey to examine how well various models perform at predicting bird richness at sites throughout North America. They train models on data from 1982-2003 and attempt to predict data from 2004-2013. They examine the performance of six different model types, two of which are null or baseline models.*

*I find it a bit surprising that the authors think that any of these richness models could predict a high level of accuracy for theses ten years. I think of richness estimations as a very coarse technique for looking at broad shifts in richness over relatively long time periods rather than these 5-species (~5-7%) shifts in richness on annual bases indicated in Figure 2. The authors don’t provide a lot of their raw data on sites other than as examples, but my impression is that they are pursuing an unrealistic level of accuracy and detail. That being said, given what the paper does, it’s okay. The models seem to be performed well. The authors give all of the keywords to indicate that they know what they are doing.*

>We agree that the relative shortness of the forecasting window and the lack of major directional shifts in richness (on average) are important to consider in interpretting our results. This is why we included long paragraphs on both topics in the initial submission. To further emphasize this point and to highlight the importance of longer time scales analyses and assessments on data showing larger changes in richness we have added an additional paragraph to the discussion addressing these points directly and recommending them for future research.

@@

*Here are some comments:*

*Birds are not the best system for doing any sort of distribution analyses. They are incredibly vagile, hard to detect, and the distributions of breeding birds are affected by many components besides environmental variables (particularly interspecific interactions). No assessment of detectability is included here other than accounting for observer identity. However, I do realize that BBS data are detailed and compelling data to use given what a large amount of detailed time-series it provides.*

>As the reviewer notes the North American Breeding Bird Survey data is one of the largest scale most long running spatiotemporal datasets available. We chose it because these features make this kind of quantitative assessment of both spatial and temporally focused forecasting methods possible. While there are downsides to any dataset we do not agree that birds uniquely problematic for this type of analysis. In addition to large amount of available data the vagility of birds potentially allows them to respond more quickly to changes in environmental conditions due to their low dispersal limitation. Imperfect detection and non-environmental drivers are common challenges in most animal species.

@@

>See comment above on hindcasting.

#### Methods:

*the authors perform their analyses on data that are collected from an incredibly dynamic time period. I wouldn’t expect species to have settled into a richness pattern given the incredibly disrupted nature of the last 32 years. Some strategies to try to handle these issues might be to train and predict at different temporal scales and to both forecast and hindcast the predictions. (e.g. train on 2004-2013 and try to predict 1982-1992 data + 1993-2003 data). If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.*

>We agree that understanding the conditions under which forecasts are likely to perform well is an important step for ecological forecasting. We have added a sentence to the discussion making this point. That said, most ecological forecasts are intended to be applied to the future, which at least at the moment we would expect to remain "incredibly dynamic". As such the dynamics of the last decade are likely a reasonable representation of the kinds of dynamics we are trying to forecast.

### 2017-12-08 manuscript/draft.Rmd
Revise Discussion in response to questions about forecastability
This addresses the concerns of both the associate editor and Reviewer 2 about
the forecastability of BBS richness in the chosen time period. In adds a new
paragraph to the Discussion that expands on our previous discussion of these
points and calls for future research in this area.

Includes associated updates to the letter.

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines that assume the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness. Ward et al.
(2014) found similar patterns in time series of fisheries stocks, where
relatively stable time series were best predicted by simple models and more
complex models were only beneficial with dynamic time series. The site-level
stability of the BBS data also explains why SDMs and macroecological models
perform relatively well at predicting future richness, despite failing to
capture changes in richness over time. However, this stability also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. These results suggest that single-site models should be actively
considered for forecasts of richness and other stable aspects of biodiversity.
Our results also suggest that future efforts to understand and forecast
biodiversity should incorporate species composition, since lower-level processes
are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
useful information [@harris2015].

>>>>

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines that assume the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness. Ward et al.
(2014) found similar patterns in time series of fisheries stocks, where
relatively stable time series were best predicted by simple models and more
complex models were only beneficial with dynamic time series. The site-level
stability of the BBS data also explains why SDMs and macroecological models
perform relatively well at predicting future richness, despite failing to
capture changes in richness over time.

The relatively stable nature of the BBS richness time-series also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. It is possible that in systems exhibiting directional changes in
richness and other biodiversity measures that models based on spatial patterns
may yield better forecasts. Future research in this area should determine if
regions or time periods exhibiting strong directional changes in biodiveristy
are better predicted by these models and also extend our forecast horizon
analyses to longer timescales where possible. Our results also suggest that
future efforts to understand and forecast biodiversity should incorporate
species composition, since lower-level processes are expected to be more dynamic
[@ernest2001; @dornelas2014] and contain more information about how the systems
are changing [@harris2015]. More generally, determining the forecastability of
different aspects of ecological systems under different conditions is an
important next step for the future of ecological forecasting.
